A Spiking Independent Accumulator Model for Winner-Take-All

### 2017-01-24 doc/paper/paper.tex
Rewriting for new results.
I still need to include the new plots. Also the paragraph about
timing still needs some adjustments.

We implement Equation~\ref{eqn:usher-mcclelland} with the NEF by using one population of neurons for each $x_i$, and applying Principle~3 to each population.
By appropriately selecting the gain and bias parameters from Principle~1, we ensure that each state variable is rectified ($x_i \ge 0$) as required.
We believe this implementation is novel as it does not interpret each $x_i$ as a distinct neural firing rate, but rather as a population-level representation distributed across any number of spiking neurons.
In effect, heterogeneous spike trains are weighted by optimal decoding weights to precisely implement the stated dynamics.
%As such, the model equations do not impose restrictions on the neuron model or requires to lump the neurons into a single population firing rate.
This allows us to attain greater biological realism without altering the dynamics prescribed by Equation~\ref{eqn:usher-mcclelland}.
\todo{distributing the states across neurons would be yet another implementation; maybe worth mentioning in the discussion...as that might be even more bio-plausible}

@@

The first layer consists of a separate integrator population (i.e., accumulator) for each state variable $x_i(t), \ i < D$.
The second layer consists of independent, non-recurrent populations that receive input from the first layer in a one-to-one fashion.
From the second layer, we decode the function $\bar{x}_i := \Theta(x_i - \vartheta)$ where $\Theta$ is the Heaviside function, and $\vartheta = 0.8$ is a fixed threshold value that determines how much evidence needs to be accumulated to produce an output.
The Heaviside decoded output of layer 2 projects back to layer 1 to add $\bar{x}_i - \bar{\beta} \sum_{j \neq i} \bar{x}_j$ to the input of $x_i$.
Since intuitively the largest input will accumulate the fastest, once this reaches the threshold $\vartheta$ it will self-excite and inhibit all other state variables.
Fixing $\bar{\beta} = 2$ ensures that the losing state variables will go to zero (see supplementary analysis).
This is summarized more precisely by the following dynamics (see Fig.~\ref{fig:indacc}):
\begin{equation}
    \begin{split}
        \frac{{\partial x}_i}{\partial t} = \rho_i \frac{1}{\tau_1} + \left( 
            \bar{x}_i - \bar{\beta} \sum_{j \neq i} \bar{x}_j \right) \frac{1}{\tau_2} , \quad x_i &\ge 0 .
    \end{split}
\end{equation}
% should we say anything about how the thresholding population works?

@@

To evaluate the two mechanisms on the previously defined input, we use a number of separate metrics.
First, we determine whether the model is able to form a clear decision within one second.
To be counted as `clear', at least one output must remain above 0.15 across the time interval $(\SI{1}{\second}, \SI{2}{\second}]$. % chktex 9
This lower bound of 0.15 was chosen to ensure that noise on a zero output is not mistaken for a non-zero output.
Note that this metric requires that the decision does not change throughout the time interval.
This does not take into account whether the winning output actually corresponds to the strongest input, but for some models it is more desirable to produce a clear incorrect decision than an unstable incorrect decision.
In other situations, though, the correctness of the decision may be of higher importance.
Thus, we consider a trial `correct' if the model forms a clear decision, and the largest output corresponds to the true largest input.

We now describe a number of metrics that will be calculated on the subset of trials that contain a clear decision.
% The time it takes to form a decision can be relevant too.
It is important to consider the speed at which the network can make decisions.
We therefore define the `decision time' as the length of time it takes to fulfil the conditions of a clear decision.
%These metrics cover the basic performance of the WTA mechanisms.
Likewise, when using a WTA mechanism within larger cognitive models, the exact magnitude of the outputs can be important.
Ideally, the output would be one for the winner and zero for all other outputs.\footnote{Technically, the winner output could be scaled to any magnitude.
We use $1$ for convenience.}
\todo{Have you tried decoding a threshold (say $\Theta(x_i - 0.15)$) off the LCA network?}
The LCA dynamics, however, prescribe an output of $u$ for the winner, with $u$ being the largest input.
We define the `winner error' $E_1$ as the difference of the winner output from $u$ for the LCA network, as the difference from one for the IA network and, for both models, the `runner-up error' $E_2$ as the difference of second largest output from zero (with negative values treated as zero).
We use the average output values over the time interval $(\SI{1}{\second}, \SI{2}{\second}]$ to calculate these error values (since these values fluctuate over time).  % chktex 9
These two metrics only consider the final output averaged over time, but the model can produce transient outputs before the final decision is achieved.
Within the context of a larger model, this can be a problem as the transient output could be prematurely interpreted as a decision.
Thus, we define $E_{2,\max}$ as the `highest output of a losing choice' during the whole simulation.

\section{Results}
We find that the ability to reach a clear decision depends on the magnitude of the highest input for both networks.
This is not surprising.
In the LCA network the output magnitude will be determined by the input magnitude of the largest input.
As such the largest input needs to be above 0.15 to fulfil the requirements.
For the IA network, the input has to be large enough to reach the integration threshold within the time limit.
By increasing the input scaling $\lambda$, or the time limit, one could get a clear decision for smaller inputs within the time limit.
We further find that the IA network is independent of the input noise and target separation for all tested parameters.\todo{but how does it do? can you put a single line on figure 3 to show it, so we can see the comparison to LCA? something visual would be good.}
This is not the case for the LCA network where the fraction of trials without a clear decision increases with noise and smaller target separation (see Fig.~\ref{fig:decisions}).
\begin{figure}
    \centering
    \includegraphics{figures/decisions}
    \caption{
        Fraction of trials with a clear decision for different input noise standard deviations $\sigma$ and target separations $s$ for the LCA network.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line shows the optimum.
    }\label{fig:decisions}
\end{figure}

However, when considering the correctness of the decisions (see Fig.~\ref{fig:correct}) the LCA network does better with the default parameters.
We can alleviate that difference to some degree by reducing the input scaling to $\lambda = 0.2$, but the IA network still performs worse for some combinations of input noise and target separation when staying below the time limit.
This is part due in part to a systematic bias for certain choices introduced in the second layer of the IA network.
In the NEF the neuron gain and bias parameters are usually picked from random distributions.
As such the layer 2 populations will become active at slightly different thresholds.
By using identical layer 2 populations, i.e.~identical gain and bias parameters for each population, this bias can be eliminated and the IA network performs better than the LCA network in most instances.
\todo{label fig 4 with a, b,c d and refer to those letters in the discussion above}
\begin{figure*}
    \centering
    \includegraphics{figures/correct}
    \caption{
        Fraction of correct trials for different input noise standard deviations $\sigma$, target separations $s$, and models.  
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line shows the optimum.
    }\label{fig:correct}
\end{figure*}

The time required to reach a decision in the LCA network depends mostly on the target separation and the amount of noise on the input (see Fig.~\ref{fig:time}\todo{put a and b here too and reference them in the discussion}), the magnitude of the largest input is of minor influence.
Also, the confidence intervals indicate a larger variance in the decision times with less target separation and more input noise.
In contrast to that, in the IA network the scaled magnitude of the largest input is the most important factor.
Depending on this magnitude the network can either be faster or slower than the LCA network, but it will need more time to achieve the same fraction of correct responses.
There is also a slight influence of target separation and input noise with an interaction of these two parameters.
\begin{figure*}
    \centering
    \includegraphics{figures/time}
    \caption{
        Mean decision times for different input noise standard deviations $\sigma$ and target separations $s$ for the LCA and IA networks.
        For the IA network $u = 0.2$ is shown which shows the interaction of target separation and input noise most clearly.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal lines show the optimum.
    }\label{fig:time}
\end{figure*}

The output of the IA network matches almost perfectly the specifications of producing an output of one for the winner and zero for all other choices \todo{a line on the LCA graph would help here too}.
This is not the case for the LCA network (see Fig.~\ref{fig:error}\todo{again a and b}).
For small target separations, the output of the winner will stay below $u$ and this will get worse with increased noise.
Furthermore, the LCA network will have residual outputs of losing choices under noisy conditions and with small target separations.
\begin{figure*}
    \centering
    \includegraphics{figures/error}
    \caption{
        Error on outputs for different input noise standard deviations $\sigma$ and target separations $s$ for the LCA network.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal lines show the optimum.
    }\label{fig:error}
\end{figure*}

@@

\section{Discussion}
We have shown that neither network performs better on all benchmarks, but rather each is better suited for different purposes.
For instance, the LCA network determines the correct winner more quickly.
However, under noisy conditions it may not produce a clear output at all, and thus fails to make a decision.
The IA network might not be as quick or find the correct winner as often, but given enough time it will eventually arrive at a decision and produce a clear output.

@@

When integrating these networks into larger neural or cognitive models, the two models present different challenges.
This integration is simpler for the IA network because the IA network provides a clear output.
The output of the winning choice is also always close to one, independent of the input, which makes it easy to detect when the decision has been made and processing can continue.
In contrast, detecting a decision is more problematic in the LCA network, where  a fluctuating output is often generated before the final decision, and the magnitude of the output depends on the input magnitude.
This makes it problematic to use a fixed threshold to detect the decision.
Furthermore, transient responses of losing choices can get close to the final output of the winning choice in the LCA network, making detection more difficult.

@@

@@

We did not investigate the biological plausibility of these mechanisms in detail because this has been done before for different WTA networks \todo{ref???}.
Also, the implementation in spiking neurons ensures a certain level of biological plausibility.
We also did not look at at the influence of the number of dimensions $D$ in detail, but qualitatively the results will stay the same.
For higher $D$, we expect less accuracy since each additional choice has a baseline chance to win due to noise.
\todo{looking only at cases where a decision could be made within the results, include a number of parameters and github url}

>>>>

We implement Equation~\ref{eqn:usher-mcclelland} with the NEF by using one population of neurons for each $x_i$, and applying Principle~3 to each population.
By appropriately selecting the gain and bias parameters from Principle~1, we ensure that each state variable is rectified ($x_i \ge 0$) as required.
We believe this implementation is novel as it does not interpret each $x_i$ as a distinct neural firing rate, but rather as a population-level representation distributed across any number of spiking neurons.
In effect, heterogeneous spike trains are weighted by optimal decoding weights to precisely implement the stated dynamics.
%As such, the model equations do not impose restrictions on the neuron model or requires to lump the neurons into a single population firing rate.
This allows us to attain greater biological realism without altering the dynamics prescribed by Equation~\ref{eqn:usher-mcclelland}.

@@

The first layer consists of a separate integrator population (i.e., accumulator) for each state variable $x_i(t), \ i < D$.
The second layer consists of independent, non-recurrent populations that receive input from the first layer in a one-to-one fashion.
From the second layer, we decode the function $\bar{x}_i := \Theta(x_i - \vartheta)$ where $\Theta$ is the Heaviside function, and $\vartheta = 0.8$ is a fixed threshold value that determines how much evidence needs to be accumulated to produce an output.
The Heaviside decoded output of layer 2 projects back to layer 1 to add $\bar{x}_i - \bar{\beta} \sum_{j \neq i} \bar{x}_j$ to the input of $x_i$.
Since intuitively the largest input will accumulate the fastest, once this reaches the threshold $\vartheta$ it will self-excite and inhibit all other state variables.
Fixing $\bar{\beta} = 2$ ensures that the losing state variables will go to zero (see supplementary analysis).
This is summarized more precisely by the following dynamics (see Fig.~\ref{fig:indacc}):
\begin{equation}
    \begin{split}
        \frac{{\partial x}_i}{\partial t} = \rho_i \frac{1}{\tau_1} + \left( 
            \bar{x}_i - \bar{\beta} \sum_{j \neq i} \bar{x}_j \right) \frac{1}{\tau_2} , \quad x_i &\ge 0 .
    \end{split}
\end{equation}
Compared to the LCA model, the threshold $\vartheta$ is an additional free parameter that controls how quickly the IA model makes a decision and how much evidence will be integrated.
Instead of directly manipulating $\vartheta$, we opt to change $\tau_1$ instead which is equivalent due to the leak-free integration.
% should we say anything about how the thresholding population works?

@@

To evaluate the two mechanisms on the previously defined input, we use a number of separate metrics.
First, we determine whether the model is able to form a clear decision within one second.
To be counted as `clear', at least one output must remain above 0.15 across the time interval $(\SI{1}{\second}, \SI{2}{\second}]$ while all other outputs remain below this threshold. % chktex 9
This lower bound of 0.15 was chosen to ensure that noise on a zero output is not mistaken for a non-zero output.
Note that this metric requires that the decision does not change throughout the time interval.
This does not take into account whether the winning output actually corresponds to the strongest input, but for some models it is more desirable to produce a clear incorrect decision than an unstable incorrect decision.
In other situations, though, the correctness of the decision may be of higher importance.
Thus, we consider a trial `correct' if the model forms a clear decision, and the largest output corresponds to the true largest input.

We use two additional benchmarks on the set of all trials with a clear decision.
First, it is important to consider the speed at which the network can make decisions.
We therefore define the `decision time' as the length of time it takes to fulfil the conditions of a clear decision.
Second, the correctness metric only considers the final averaged output during a time interval.
It is possible for a network to produce transient outputs before the final decision is reached.
In the context of a larger model, this can become a problem because the transient output might be prematurely interpreted as a decision.
Thus, we define $E_{\max}$ as the `highest output of a losing choice' during the whole simulation.

\section{Results}
We find that the ability to form a clear decision of the LCA network decreases with more noise and less target separation (see Fig.~\ref{fig:decisions}).
The best performance is achieved with a medium baseline of $u=0.6$.
A higher baseline makes it more likely that runner-ups will exceed the 0.15 threshold, especially wit ha small target separation.
With a small baseline, on the other hand, even the winner can be unable to exceed the threshold in the presence of noise.
In contrast to the LCA network, the IA network manages to form a clear decision in every trial (not explicitly shown in Figure~\ref{fig:decisions}, but all data points fall on the grey line).
\begin{figure}
    \centering
    \includegraphics{figures/decisions}
    \caption{
        Fraction of trials with a clear decision for different input noise standard deviations $\sigma$ and target separations $s$ for the LCA network.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line shows the optimum.
    }\label{fig:decisions}
\end{figure}

Interestingly, in 99.97\% of successful decisions, the correct winner was determined by the LCA network.
Thus, a plot of correct trials looks virtually identical to Figure~\ref{fig:decisions}.
While always reaching a decision, the decisions of the IA network are not always correct (see Fig.~\ref{fig:correct}).
Overall, the IA performance tends to worse for a high baseline, but better for a low baseline.
We can greatly improve the IA performance by increasing $\tau_1$ which effectively adjusts the threshold $\vartheta$.
This will, however, make the network slower to decide.
A further improvement is possible by eliminating a systematic bias introduced in the second layer of the IA network.
In the NEF the neuron gain and bias parameters are usually picked from random distributions.
As such the layer 2 populations will become active at slightly different thresholds.
By using identical layer 2 populations, i.e.~identical gain and bias parameters for each population, this bias can be eliminated and the IA network performs even better.
\todo{label fig 4 with a, b,c d and refer to those letters in the discussion above}
\begin{figure*}
    \centering
    \includegraphics{figures/correct}
    \caption{
        Fraction of correct trials for different input noise standard deviations $\sigma$, target separations $s$, and models.  
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line shows the optimum.
    }\label{fig:correct}
\end{figure*}

The time required to reach a decision in the LCA network depends mostly on the baseline and target separation (see Fig.~\ref{fig:time}\todo{put a and b here too and reference them in the discussion}).
Similarly, in the IA network the magnitude of the largest input is also the most important factor.
Depending on this magnitude, the network can either be faster or slower than the LCA network, but it will need more time to achieve the same fraction of correct responses.
There is also a slight influence of target separation and input noise with an interaction of these two parameters.
\todo{this might need some rewriting given new results}
\begin{figure*}
    \centering
    \includegraphics{figures/time}
    \caption{
        Mean decision times for different input noise standard deviations $\sigma$ and target separations $s$ for the LCA and IA networks.
        For the IA network $u = 0.2$ is shown which shows the interaction of target separation and input noise most clearly.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal lines show the optimum.
    }\label{fig:time}
\end{figure*}

@@

\section{Discussion}
We have shown that neither network performs better on all benchmarks, but rather each is better suited for different purposes.
For instance, the LCA network determines the correct winner more quickly and given a conclusive decision was made it is almost always the correct decision.
However, under noisy conditions it may fail to produce a clear output at all, and thus not make a decision.
This can be problematic in situations were it is better or required to make a decision even though it might not be the best choice.
The IA network might not be as quick or find the correct winner as often depending on the choice of $\tau_1$, but given enough time it will eventually arrive at a decision and produce a clear output.

@@

Both models might produce a transient response that can be problematic in detecting when the decision has been made as it might be interpreted as the decision.
This is of special relevance when integrating the networks in larger cognitive models.
For the LCA network this behaviour is inherent.
There will always be an initial rise of all state variables that receive input before the mutual inhibition grows strong enough to push them back to zero.
In the IA network, however, this is mostly a matter of choosing an appropriate $\tau_1$.

@@

Relatedly, it might be more plausible to distribute the representation of all state variables over a single population of neurons.
This is directly supported by the NEF, but we chose to leave this to future work to keep the current analysis free from potential interactions of the state variables introduced by such a distributed representation.

It is worth to highlight, that one can transform one model into the other with a few steps.
By setting $\tau = \tau_1$, $k = -\tau/\tau_2$, and $\beta = \bar{\beta}\tau/\tau_2$ we obtain almost identical model equations.
The only remaining difference is the added Heaviside non-linearity on the state variables.
Nevertheless, the behaviour of the two models is quite distinct.

@@

We did not investigate the biological plausibility of these mechanisms in detail because this has been done before for different WTA networks \todo{ref???}.
Also, the implementation in spiking neurons ensures a certain level of biological plausibility.
We also did not look at at the influence of the number of dimensions $D$ in detail, but qualitatively the results will stay the same.
For higher $D$, we expect less accuracy since each additional choice has a baseline chance to win due to noise.
